---
title: "BOSTON HOUSING MARKET PREDICTION"
author: "steve dubois"
date: "4/4/2019"
output:
  html_document: default
  pdf_document: default
---
<style type="text/css">

h1.title {
  font-size: 38px;
  color: DarkRed;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}
</style>

#Overview
In this report I detail the machine learning (ML) models I implemented to accurately predict the housing prices in Boston suburbs. The data set for this experiment is accessed from the UCI Machine Learning repository via https://archive.ics.uci.edu/ml/datasets/Housing. The report is organized in such a way as to demonstrate the entire process right from getting and cleaning the data, to exploratory analysis of the data set to understand the distribution and importance of various features in influencing the algorithm, to coming with a hypothesis, training ML models, evaluation of the models, etc.


##Introduction
The data set consists of 506 observations of 14 attributes. The median value of house prices in $10000s, denoted by MEDV, is the outcome or the dependent variable in our models.   As our goal is to develop a model that has the capacity of predicting the price/value of houses, we will split the dataset into features and the target variable.  Along the way, we do also discover which neighborhood attibutes are most significant to housing price prediction.

The features ‘RM’, ‘LSTAT’ and ‘PTRATIO’, give us quantitative information about each datapoint. We will store them in features.
The target variable, ‘MEDV’, will be the variable we seek to predict. We will store it in prices.

###brief description of each predictor feature and the outcome in our data set:
variables:

CRIM – per capita crime rate by town"
ZN – proportion of residential land zoned for lots over 25,000 sq.ft"
CHAS – Charles River dummy variable (1 if tract bounds river; else 0)" 
NOX – nitric oxides concentration (parts per 10 million)"
RM – average number of rooms per dwelling"
AGE – proportion of owner-occupied units built prior to 1940"
DIS – weighted distances to five Boston employment centers"
RAD – index of accessibility to radial highways"
INDUS – proportion of non-retail business acres per town"
TAX – full-value property-tax rate per $10,000"
PTRATIO – pupil-teacher ratio by town"
B – 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town"
LSTAT – % lower status of the population"
MEDV - Median value of owner-occupied homes in $10000’s (response variable"




##LOADING REQUISITE R PACKAGES
```{r, suppressMessages = TRUE, echo=FALSE, warning = FALSE, message=FALSE,tidy=TRUE}
library(knitr)
library(broom)
library(tidyverse)
library(tidyr)
library(dplyr)
library(readr)
library(ggplot2)
library(randomForest)
library(caret)
library(data.table)
library(ggplot2)
library(plotly)
library(rattle)
library(GGally)
library(ggfortify)
```



##GETTING DATA
Getting the data into R as an R object, cleaning the data and transforming it as a neat and usable R data frame or equivalent. The df (Boston housing) file consists of the actual data,  The R-function readLiness() reads data from fixed-width files. Below, I use subsetting and the strsplit R function to extract the columns/predictors and column names alone from this file.
```{r, echo=TRUE, message=FALSE,tidy=TRUE, warning=FALSE}
# DATA DOWNLOADED
text <- readLines("boston.txt")
text <- text[c(-1, -2)]
```

##PREPROCESSING/TRANSFORM DATA
```{r, echo=TRUE, message=FALSE,tidy=TRUE}
i = 1
df2 <- NULL
while (i <= 1012) {
    if (i %% 2 == 0) {
    i = i + 1 } else i 
    j <- i + 1
texti <- as.numeric(strsplit(text, " ")[[i]])
texti <- na.omit(texti)
textj <- as.numeric(strsplit(text, " ")[[j]])
textj <- na.omit(textj)
textC <- as.vector(c(texti, textj))
df <- NULL
df <- rbind(df2, textC)
colnames(df) <- c("CRIM", "ZN", "INDUS", "CHAS", "NOX", "RM", "AGE", "DIS",
"RAD", "TAX", "PTRATIO", "B", "LSTAT", "MEDV") 
rownames(df) <- c()
df2 <- df
i <- j + 1
df <- as.data.frame(df) }
```

###first 10 observations
```{r, echo=TRUE, message=FALSE,tidy=TRUE}
head(df, 7)
```

#EXPORATORY DATA ANALYSIS
Now, let us check and explore the cleaned data frame containing the housing data. The df R object is of class ‘data.frame’, which is very easy to work with using R scripts. The str() function is powerful in displaying the structure of an R dataframe. Below, the output of str() compactly provides the relevant information of our dataframe, like the number of observations, number of variables, names of each column, the class of each column, and sample values from each column.


##DiSPLAY SUMMARY STATISTICS OF DATA
```{r, echo=TRUE, message=FALSE,tidy=TRUE}
summary(df)
```


##DiSPLAY SUMMARY STRUCTURE OF HOUSING DATA (df)
```{r, echo=TRUE, message=FALSE,tidy=TRUE}
str(df)
```


##CORELLATION & NEAR ZERO VALUE
A few important properties to check now are the correlation of input features with the dependent variable, and to check if any feature has near zero variance (values not varying much within the column).
```{r, echo=TRUE, message = FALSE,tidy = TRUE}
```
##STRENGTH OF CORRELATION HOUSING ATTRIBUTE TO HOUSING PRICES - "MEDV"
```{r, echo=TRUE, message=FALSE,tidy=TRUE}
suppressMessages(library(caret))
pp <- cor(df, df$MEDV) 
pp1 <- cbind(rownames(pp), pp)
pp1 <- as.data.frame(pp1)
colnames(pp1) <-  c("Atribute_Feature", "Correlation_to_Prices")
pp1 <- pp1 %>% arrange(desc(Correlation_to_Prices))
pp1

p06 <- ggplot(pp1) +  geom_col(aes(x = Atribute_Feature, y = Correlation_to_Prices, fill = Correlation_to_Prices)) + ggtitle("STRENGTH OF CORRELATION HOUSING ATTRIBUTE TO HOUSING") 

ggplotly(p06 = ggplot2::last_plot()) %>% layout(width = 1000, height = 500)
```
##CORRELATION HOUSING ATTRIBUTE TO HOUSING
####Calulate near zero variance
```{r, echo=TRUE, message = FALSE,tidy = TRUE}
nzv <- nearZeroVar(df, saveMetrics = TRUE)
sum(nzv$nzv)
```
  

#EXPLORATORY PLOTS
Let us visualize the distribution and density of the outcome, MEDV. The black curve represents the density. In addition, the boxplot is also plotted to bring an additional perspective. We see that the median value of housing price is skewed to the right, with a number of outliers to the right. It may be useful to transform ‘MEDV’ column using functions like natural logarithm, while modeling the hypothesis for regression analysis.

```{r, echo=TRUE, message=FALSE,tidy=TRUE}
library(corrplot)
p01 <- ggplot(df, aes(x = MEDV)) + xlab("Median value of owner-occupied homes")  + geom_histogram(aes(y = ..density..), binwidth = 1, fill = "green") + geom_density(alpha = 0.8, fill = "red") + ylim(0, 0.075)
ggplotly(p01 = ggplot2::last_plot())
           
p02 <- ggplot(df, aes(y = MEDV, x = RM)) + xlab("average number of rooms per dwelling") + ylab("Median value of owner-occupied homes") + geom_point(colour = "red") + geom_smooth(method = lm, colour = "blue") + geom_smooth(colour = "orange")
ggplotly(p02 = ggplot2::last_plot())

p03 <- ggplot(df, aes(y = MEDV, x = LSTAT)) + xlab("%lower status of the population") + ylab("Median value of owner-occupied homes") + geom_point(colour = "green") + geom_smooth(method = lm, colour = "yellow") + geom_smooth(colour = "red")
ggplotly(p03 = ggplot2::last_plot())

p04 <- ggplot(df, aes(y = MEDV, x = PTRATIO)) + xlab("pupil-teacher ratio by town") + ylab("Median value of owner-occupied homes") + geom_point(colour = "black") + geom_smooth(method = lm, colour = "yellow") + geom_smooth(colour = "green")
ggplotly(p04 = ggplot2::last_plot())

p05 <- ggplot(df, aes(y = MEDV, x = CRIM)) + xlab("per capita crime rate by town") + ylab("Median value of owner-occupied homes") + geom_point(colour = "black") + geom_smooth(method = lm, colour = "yellow") + geom_smooth(colour = "red") 

ggplotly(p05 = ggplot2::last_plot()) %>% layout(autosize = F, width = 500, height = 500)



```

```{r, echo=TRUE, message=FALSE,tidy=TRUE} 
b0x <- ggplot(df, aes(y = MEDV, x = c(1:506))) + geom_boxplot(alpha = 0.8, colour = "black", fill = "green") + coord_flip()
b0x



```


Now, let us do scatter plot of some of the important variables (based on intuition) with the outcome variable MEDV. We see that there is strong positive or negative correlation between these variables and the outcome. It is also obviously evident that INDUS and NOX are strongly positively correlated with one another, as nitric oxide levels tend to go up with increase in industries.
```{r, echo = TRUE}
plot(df[,c(3,5,6,11,13,14)],pch = 3)
```
We see that the number of rooms RM has the strongest positive correlation with the median value of the housing price, while the percentage of lower status population, LSTAT and the pupil-teacher ratio, PTRATIO, have strong negative correlation. The feature with the least correlation to MEDV is the proximity to Charles River, CHAS.



##Feature Engineering and Data Partitioning
Next we perform centering and scaling on the input features. Then we partition the data on a 7/3 ratio as training/test data sets.
```{r, echo = TRUE, supppressMessages = TRUE}
 #enteringcaling of input features
df <- cbind(scale(df[1:13]), df[14])
```

###Setting seed to maintain reproducibity 
```
set.seed(12345)
`````

## CREATE TEST & TRAIN DATA PARTITIONS 
```{r, echo = TRUE, supppressMessages =  TRUE}
inTrain <- createDataPartition(y = df$MEDV, p = 0.70, list = FALSE)
training <- df[inTrain,]
testing <- df[-inTrain,]
```



##DEVELOPING A MODEL FOR PREDICTION
I'm going to develop the tools and techniques necessary for a model to make a prediction. Being able to make accurate evaluations of each model's performance helps to greatly reinforce the confidence in my predictions.

##Defining a Performance Metric
###R2 (R-squared) -- "Coefficient of determination.  
The values for R2 range from 0 to 1, which captures the percentage of squared correlation between the predicted and actual values of the target variable. A model with an R2 of 0 is no better than a model that always predicts the mean of the target variable, whereas a model with an R2 of 1 perfectly predicts the target variable. Any value between 0 and 1 indicates what percentage of the target variable, using this model, can be explained by the features. A model can be given a negative R2 as well, which indicates that the model is arbitrarily worse than one that always predicts the mean of the target variable.


###RMSE (Root Mean Square Error)
is also a good measure of how accurately the model predicts the response, and it is the most important criterion for fit if the main purpose of the model is prediction.  Similar to R2, its values ramge from 0 to 1.  However, the differene from R2 is that closer RMSE is 0, indicates a better fit to regression, implying better prediction accuracy. 

##LINEAR REGRESSIONS

First, let us try generalized linear regression model with MEDV as the dependent variable and all the remaining variables as independent variables. We train the model with the training data set. For this linear model, below are the coefficients of all the features, and the intercept. Next, we use the trained model to predict the outcome (MEDV) for the testing data set. A good metric to test the accuracy of the model is to calculate the root-mean squared error, which is given by :

$$\sqrt{\sum_{i=1}^{n} \frac{(y_{pred_i} - y_{act_i})^2} {n}}$$
 
 ##LINEAR REGRESSION MODEL 1 - using all features
```{r, echo=TRUE, message=FALSE,tidy=TRUE}
set.seed(12345)
fit.lm <- lm(MEDV~.,data = training)

fit.lm
autoplot(fit.lm,  which = 1:6, ncol = 3, label.size = 3)


####CHECK COEFFICIENTS

data.frame(coef = round(fit.lm$coefficients,2))
summary(fit.lm)[12]


ggplot(data = df, aes(df$NOX, df$MEDV)) + geom_smooth()
```


```{r, echo=TRUE, message=FALSE,tidy=TRUE}
set.seed(12345)

#predict on test set
pred.lm <- predict(fit.lm, newdata = testing)

#Root-mean squared error
rmse.lm <- sqrt(sum((pred.lm - testing$MEDV)^2)/length(testing$MEDV))
                   
c(RMSE = rmse.lm, R2 = summary(fit.lm)$r.squared)
  summary(fit.lm)$coefficients[2,4]
data.frame(RMSE = rmse.lm, R2 = summary(fit.lm)[9], p.value = summary(fit.lm)$coefficients[2,4])



```

We see that the RMSE is 4.381992 and the R2R2 value is 0.7239 for this model.


###LINEAR REGRESSION MODEL 2

We also saw that the output variable MEDV was skewed to the right. Performing a log transformation would normalize the distribution of MEDV. Let us perform glm with log(MEDV) as the outcome and all remaining features as input. We see that the RMSE value has reduced for this model.

###Try linear model using significant features
```{r, echo=TRUE, message=FALSE,tidy=TRUE}
set.seed(12345)

fit.lm1 <- glm(log(MEDV) ~ CRIM + CHAS + NOX + RM + DIS + PTRATIO + RAD + B + LSTAT, data = training)


#predict on test set
summary(fit.lm1)

#predict on test set
pred.lm1 <- predict(fit.lm1, newdata = testing)
pred.lm1 <- pred.lm1*100000
range(pred.lm1)

#Root-mean squared error
rmse.lm1 <- sqrt(sum((exp(pred.lm1) - testing$MEDV)^2)/length(testing$MEDV))

c(RMSE = rmse.lm1, R2 = summary(fit.lm1)$r.squared, P_value = summary(fit.lm1)$coefficients[1,4])

c(RMSE = rmse.lm1/10, R2 = summary(fit.lm1)$r.squared)


```
We see that the RMSE is 4.381992 and the R2 value is 0.7239 for this model.



Let us examine the calculated p-value for each feature in the linear model. Any feature which is not significant (p<0.05) is not contributing significantly for the model, probably due to multicollinearity among other features. We see that the features, ZN, INDUS, and AGE are not significant.
```{r, echo=TRUE, message=FALSE,tidy=TRUE}```
vif(fit.lm1)
```
Variance inflation factors are computed using vif() for the standard errors of linear model coefficient estimates. It is imperative for the vif to be less than 5 for all the features. We see that the vif is greater than 5 for RAD and TAX.


log(MEDV) ~ CRIM + CHAS + NOX + RM + DIS + PTRATIO + RAD + B + LSTAT


```{r, echo=TRUE, message=FALSE,tidy=TRUE}```

This model is marginally less accurate than linear model 2, based on slight increase in RMSE and slight decrease in R2R2 value. Let us plot the predicted vs actual values of the outcome MEDV.

## LINEAR MODEL 2 PLOT of Predicted Prices vs Actual Prices
```{r, echo=TRUE, message=FALSE,tidy=TRUE}
plot(pred.lm1,testing$MEDV, xlab = "Predicted Price", ylab = "Actual Price")

#diagnostics plots
layout(matrix(c(1,2,3,4),2,2))
plot(fit.lm1)
```

@@#LINEAR MODEL 2 -- TABLE
Table Shoming 1st six observations – Actual vs Predicted Price

```{r, echo=TRUE, message=FALSE,tidy=TRUE}
table <- data.frame(x = pred.lm1, y = testing$MEDV)
names(table) <- c(xlab = "Predicted_Price", ylab = "Actual_Price")
data.table(table,6)


```

#RANDOM FOREST MODEL

For random forest implementation, we could use the linear model formula of MEDV ~ . (meaning MEDV is the outcome with all other features as input). Inspecting the results, we see that the random forest model has given the best accuracy so far.  Better model peformance is expected.
```{r, echo=TRUE, message=FALSE,tidy=TRUE}

library(randomForest)
set.seed(12345)
fit.rf <- randomForest(MEDV ~ ., data = training) 
fit.rf

pred.rf <- predict(fit.rf, testing)


rmse.rf <- sqrt(sum(((pred.rf) - testing$MEDV)^2)/length(testing$MEDV))
```

```{r, echo=TRUE, message=FALSE,tidy=TRUE}
table <- c(RMSE = rmse.rf, Rsquared = mean(fit.rf$rsq)) 
print(table)
```

##RANDOM FOREST PLOT -- Predicted Prices vs Actual Prices
```{r, echo=TRUE, message=FALSE,tidy=TRUE}
plot(pred.rf, testing$MEDV, xlab = "Predicted Price", ylab = "Actual Price", pch = 3)

varImpPlot(fit.rf,main = "Important variables for Yield 2015", fill = "red")

```

##RANDOM FOREST MODEL TABLE 
Table Shoming 1st TEN observations --  Actual vs Predicted Price 
```{r, echo=TRUE, message=FALSE,tidy=TRUE}
table1 <- data.frame(x = pred.rf, y = testing$MEDV)
names(table1) <- c("Predicted_Price", ylab = "Actual_Price")
head(kable(table1),10)
```


##MODEL COMPARISON
```{r, echo=TRUE, message=FALSE,tidy=TRUE}
Linear_Model_1 <- c(RMSE = rmse.lm/10, R2 = summary(fit.lm)$r.squared)
Linear_Model_2 <- c(RMSE = rmse.lm1/10, R2 = summary(fit.lm1)$r.squared)/10
Random_Forest_Model <- c(RMSE = rmse.rf, R2 = mean(fit.rf$rsq))

model_comparison <- rbind(Linear_Model_1, Linear_Model_2, Random_Forest_Model)
kable(model_comparison)

```


##CONCLUSIONS - INFERENCES
We experimented with a couple of linear regression models and a random forest model to predict the housing prices in Boston suburbs. Among these models, the Random forest model with a simple linear relationship between the outcome and all input features yielded the best model to predict outcomes, as determined from having the smallest RMSE (i.e., root mean squared error) and the highest R2 (i.e., accuracy | R-squared statistic), and the smallest p-value (greatest significance).

In addition the goal of this report was to determine the best predictive algorithymic model , its aim was also to illuminte the several neighborhood attributes that best explained variation in house pricing.  Various statistical techniques were used to eliminate predictors and extraneous observations. In examining the final model, one finds – quite reasonably – that house prices are higher in areas with lower crime and lower pupil-teacher ratios. House prices also tend to be higher closer to the Charles River, and houses with more rooms are pricier. This report is interested in the neighborhood attributes of houses, so the number of rooms is not an important predictor. The most interesting factors to consider are nitrogen oxide levels and distance to the main employment centers. On the one hand, people would want to live close to their place of employment. Yet it is reasonable to suggest that pollution levels are higher as one moves closer to these main employment centers. Most importantly, when talking of pollution, it is not just nitrogen oxide levels that are higher, but also noise pollution levels. The regression model that was fitted shows that higher levels of pollution decrease house prices to a greater extent than distance to employment centers. This suggests that people would prefer to live further away from their place of employment if it meant lower levels of pollution, which is an interesting point to consider. On a concluding note, it is important to note that the data for this report was collected several decades ago. In the years since, there is no doubt that pollution levels have risen and it would be interesting to examine the ways in which that affects house pricing in Boston today.





